---
title: "Class 11"
author: "Courtney Cameron PID:A69028599"
format: pdf
editor: visual
---

# AlphaFold 

alphafold can be run through GoogleColab: https://colab.research.google.com/github/sokrypton/ColabFold


```{r}
pth <- "hivprdimer_23119/"

list.files(path=pth)
```

The multiple sequence alignement (MSA) is conatined in the 'a3m' file 

return full a3m file name within the directory of the alignment file
```{r}
aln.file <- list.files(path=pth, pattern = '.a3m', all.files=FALSE, full.names = TRUE)
```


```{r}
library(bio3d)
aln <- read.fasta(aln.file, to.upper = TRUE)
```

```{r}
attributes(aln)
```

```{r}
dim(aln$ali)
```

Calculating sum summary info such as conservation scores
```{r}
sim <- conserv(aln)
```

```{r}
plot(sim, typ='h')
```

the coserved columns can be summarized with a consensus sequence

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```

reading in the structure models

Read the PAE (predicted aligned arror files) into R, these files are stored in 
JSON format

```{r}
library(jsonlite)
```

```{r}
pae.file <- list.files(path=pth,pattern='.000.json',full.names = TRUE)
pae.file
```

```{r}
pae5 <- read_json(pae.file[5], simplifyVector = TRUE)
pae1 <- read_json(pae.file[1], simplifyVector = TRUE)
```


```{r}
dim(pae1$pae)
```
```{r}
plot.dmat(pae1$pae)
```

```{r}
plot.dmat(pae5$pae)
```




















