---
title: "Class 6"
author: "Courtney Cameron PID:A69028599"
format: pdf
editor: source
---

#Example function
3 parts to a function:

- a name
- input (none,one,or more)
- a body

function to add two numbers

```{r silly_function}
sillyadd <- function(x,y=1) {
  x+y
}
```

```{r}
sillyadd(100,4)
```
## Q1

# Functions for Lab

Example input vectors to start with 

```{r}
student1<-c(100,100,100,100,100,100,100,90) 
student2<-c(100,NA,90,90,90,90,97,80) 
student3<-c(90,NA,NA,NA,NA,NA,NA,NA)


mean(student2)
```



```{r}
grade <- function(x) {
  x[is.na(x)] <- 0
  y <- which.min(x)
  mean(x[-y])

}
```

testing aspects of the function
```{r}
#find lowest value
which.min(student1)
#mean of list excluding lowest value
mean(student1[-8])
#changing NA to 0
student3[is.na(student3)]<-0
student3

```


running function
```{r}
grade(student1)
grade(student2)
grade(student3)
```

loading in data from csv file
```{r}
url <- 'https://tinyurl.com/gradeinput'

grade_book <- read.csv(url,row.names=1)

```

apply function to the to the gradebook 
apply(data, row or column #, function)
```{r}
finalgrade <- apply(grade_book, 1 ,grade)
finalgrade
```
## Q2
find highest scoring student
```{r}
which.max(finalgrade)
```

## Q3
finding lowest score using mean
  use na.rm=TRUE to ignore NA in the data
```{r}
low <- apply(grade_book, 2,mean, na.rm=TRUE)

low
which.min(low)
```

## Q4
Correlation of homework score to overall score

```{r}
mask <- grade_book
mask[is.na(mask)] <-0

cor(mask$hw5, finalgrade)
cor(mask$hw3, finalgrade)

apply(mask, 2, cor, y=finalgrade)
```



