---
title: "Class5"
author: "Courtney Cameron PID: A69028599"
format: pdf
---

# Graphic systems in R

Base R graph

```{r}
plot(cars)
```

ggplot2 graph

```{r}
library(ggplot2)
```

each graph needs 3 layers:

-   **data**

-    **aes**

-    **geoms**

```{r}

ggplot(cars, aes(x=speed, y=dist)) +
  geom_point()
```

```{r}
ggplot(cars, aes(x=speed, y=dist)) +
  geom_point() +
  geom_smooth() +
  labs(title = 'Stopping Distance of Old Cars') +
  theme_bw()
```

```{r}
p <- ggplot(cars, aes(x=speed, y=dist)) +
  geom_point()

p + geom_smooth(method = 'lm', se = FALSE)
```

```{r}
library(patchwork)
```

```{r}
p | p
```
#Adding more plot aestethics through aes()

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```


```{r}
nrow(genes)
colnames(genes)
ncol(genes)

table(genes$State)

round(table(genes$State)/nrow(genes)*100,2)
```

```{r}
p2<- ggplot(genes, aes(x=Condition1, y=Condition2, col = State)) +
  geom_point()

p2
```

```{r}
p2 + scale_color_manual(values = c('red','grey','blue'))
```

```{r}
p2 + scale_color_manual(values=c('red','grey','blue')) +
  labs(title= 'Gene Expression Changes Upon Drug Treatment', x= 'Control (no drug)', y='Drug Treatment')

```

#Using plotly to generate annotated graphs


```{r}
#| message: false
library(plotly)
```
```{r}
p3 <- ggplot(genes, aes(x=Condition1, y=Condition2, col = State, name = Gene)) + geom_point()+
  scale_color_manual(values=c('red','grey','blue')) +
  labs(title= 'Gene Expression Changes Upon Drug Treatment', x= 'Control (no drug)', y='Drug Treatment')




#ggplotly(p3)
```



#Going Further


```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

```{r}
library(dplyr)

gapmainder_2007 <- gapminder %>% filter(year==2007)
```

```{r}
ggplot(gapmainder_2007, aes(x=gdpPercap, y=lifeExp)) +
  geom_point(alpha=0.4)
```

```{r}
ggplot(gapmainder_2007, aes(x=gdpPercap, y=lifeExp, col=continent, size=pop)) +
  geom_point(alpha = 0.5)
```

```{r}
ggplot(gapmainder_2007, aes(x=gdpPercap, y=lifeExp, col=pop)) +
  geom_point(alpha = 0.8)
```

```{r}
ggplot(gapmainder_2007, aes(x=gdpPercap, y=lifeExp,size=pop)) +
  geom_point(alpha = 0.5)
```

```{r}
ggplot(gapmainder_2007, aes(x=gdpPercap, y=lifeExp,size=pop)) +
  geom_point(alpha = 0.5) +
  scale_size_area(max_size = 10)
```

#gapmider 1957 data

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957)

 ggplot(gapminder_1957, aes(x=gdpPercap, y=lifeExp, col=continent,size=pop)) +
  geom_point(alpha=0.7) +
  scale_size_area(max_size = 10) +
   labs(title='1957')



```

#Using Facet_wrap to combine graphs of the same data set

```{r}
 gapminder_1957_2007 <- gapminder %>% filter(year==1957|year==2007)
 
  ggplot(gapminder_1957_2007, aes(x=gdpPercap, y=lifeExp, col=continent,size=pop)) +
  geom_point(alpha=0.7) +
  scale_size_area(max_size = 10) +
    facet_wrap(~year)
```


#Bar charts 

Sort to have the top 5 based on population
```{r}
gapminder_top5 <- gapminder %>% 
  filter(year==2007) %>% 
  arrange(desc(pop)) %>% 
  top_n(5, pop)

gapminder_top5
```

```{r}
ggplot(gapminder_top5, aes(x=country, y=pop)) +
  geom_col()
```

```{r}
ggplot(gapminder_top5, aes(x=country, y=lifeExp)) +
  geom_col()

ggplot(gapminder_top5, aes(x=country, y=lifeExp, fill=continent)) +
  geom_col()


ggplot(gapminder_top5, aes(x=country, y=lifeExp, fill=lifeExp)) +
  geom_col()
```


```{r}
ggplot(gapminder_top5, aes(x=country, y=pop, fill = gdpPercap)) +
  geom_col()

ggplot(gapminder_top5, aes(x=reorder(country, -pop), y=pop, fill = gdpPercap)) +
  geom_col()

ggplot(gapminder_top5, aes(x=reorder(country, -pop), y=pop, fill=country))+
  geom_col(col='grey30') +
  guides(fill='none')
```

#Flipping bar charts

```{r}
head(USArrests)
```

```{r}
USArrests$State <- rownames((USArrests))

ggplot(USArrests, aes(x=reorder(State,Murder), y=Murder)) +
  geom_col() +
  coord_flip()
```

```{r}

ggplot(USArrests, aes(x=reorder(State,Murder), y=Murder)) +
  geom_point() +
  coord_flip()+
  geom_segment(aes(x=State, xend=State, y=0, yend=Murder,),col='blue')
```

#Extensions: Animations

graph is commented out for pdf format

```{r}
library(gapminder)
library(gganimate)

#ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = continent)) +
 # geom_point(alpha = 0.7, show.legend = FALSE) +
# scale_colour_manual(values = country_colors) +
  #scale_size(range = c(2, 12)) +
  #scale_x_log10() +
  #facet_wrap(~continent) +
  #labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  #transition_time(year) +
  #shadow_wake(wake_length = 0.1, alpha = FALSE)



```

#combining plots
```{r}
p1 <- ggplot(mtcars) + geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) + geom_boxplot(aes(gear, disp, group = gear))
p3 <- ggplot(mtcars) + geom_smooth(aes(disp, qsec))
p4 <- ggplot(mtcars) + geom_bar(aes(carb))

(p1|p2|p3)/p4

```

